---
title: "A protocol for GBIF data exploration and filtering for Integrated Species Distribution Models - with a practical example"
bibliography: bibliography.bib
format: html
editor: visual
---

# Introduction

Modern ecological research is increasingly coming to be characterized by the use of large, multi-source data sets from open-access online repositories - with the infrastructure provided by the Global Biodiversity Information Facility (GBIF) as perhaps the most prominent example. These "big data" on species occurrences are often used to inform species distribution models. However, the fact that the data resulting from a single GBIF query (or a similar repository) often come from widely different sources, with different levels of sampling structure, information on species absences etc., remains a challenge.

Recent advances in species distribution modelling, exemplified with the release of the `R` package `pointedSDM` (Mostert & O'Hara, 2023), does increasingly allow for the combination of multiple species occurrence data sets with different levels of sample information (i.e. presence-only, presence-absence, counts etc), into a single integrated model (iSDM). Still, practical and actionable advice for how to best obtain, validate and prepare this differentiated data from an online repository, seems to be comparatively lacking in the literature.

In this report, we aim to address this issue by demonstrating a suggestion for a complete and reproducible workflow for data download, cleaning, exploration and differentiating, from a simple GBIF query to fitting an integrated SDM with the `pointedSDM` package.

## Case study

As a case study, we will use the occurrences of the mayfly species *Beatis rhodani*, whose GBIF occurrence data in Norway is known to contain both structured and unstructured samples. In order to assess the value of our approach over more "naïve" ways of modelling the data, we will compare the output of four different SDMs:

-   A presence-only version of our complete GBIF dataset

-   A presence-absence model of the subset of the data that is determined to be from structured survey

-   A presence-only model of the subset of the data,

-   An informed combination of the presence-absence and presence-only subsets of the data.

This report documents the different steps in the mayfly case study, created for the course Advanced biology BI8091 at NTNU, Spring 2024. The analysis consists of several steps: Selection of invertebrate data points, preparation of environmental variables, running the SDMs and interpreting/discussing the results and possible implications.

### Species distribution modelling

The aim of the project is to investigate how species distribution models can be used with different data types (presence-only, presence-absence, occurrence), and how the inclusion of both "good" and "less informative" data impact possible future predictions of species distribution. The R package `pointedSDMs` "[https://CRAN.R-project.org/package=PointedSDMs](https://cran.r-project.org/package=PointedSDMs)" by Mostert & O'Hara (2023) is used for species distribution modelling.

### Study organism

The case study will focus on the mayfly species *Baetis rhodani* , also called the Large Dark Olive Mayfly. Nymphs of *B. rhodani* is commonly found in running waters all over Norway (Artsdatabanken, 2021) and can also be found in most of Western-Europe. *B. rhodani* adults are winged and inhabit the terrestrial habitat for a limited time in order to reproduce. They cannot feed in their adult stage, so their life on land is short - typically lasting a few days. Not many studies have measured the dispersal distance of adult Mayflies. A study from 2021 aggregating flying distances for EPTs (Peredo Arce et al., 2021) cite only one study for *B. rhodani*, which had measured adults flying more than 3700 meters (Bagge, 1995).

We assume that most of the *B. rhodani* datapoints from Norway originate from aquatic sampling, which for some datasets can be verified (ie., the NTNU University Museum).

### Data

Mayfly data is accessed through the Global Biodiversity Information Facility (GBIF, 2024). Norway is selected as study area, due to the availability of both high-quality species occurrence data (NTNU University Museum) and presence-only data from national monitoring and research. Easily accessible physical, climatic and geographical data also makes Norway a good case-study.

# Workflow

## 0. Setup

Load packages: Downloading and loading required packages using a function that checks whether each package is already downloaded, and only downloads it if not present.

```{r 0_setup}
#| output: false
source(here::here("code","0_setup.R"))
```

## 1. Data download and exploration

### 1.1. Create download request for *Baetis rhodani* data from GBIF

In this project, we use data on the mayfly *Baetis rhodani* accessed via GBIF. The GBIF infrastructure provides several options for data download, which all require a GBIF user. Creating a user is free and open for all. Searching for and downloading data can be done via the GBIF website, or it can be done via R using the `rgbif` package. We will demonstrate the latter, at it improves reproducibility.

First, a download request is created. This is shown in the script `1_1_GBIF_download.R`. Here we specify the download criteria:

-   Species: *Baetis rhodani* (Pictet, 1843)

-   Global administrative area: Norway

-   Basis of record: Everything

-   Year range: 1950-current

Then, the data is imported and saved as `insectdata.rda`.

### 1.2. GBIF download using dataset key

Creating a download request from GBIF only needs to be done once. Afterwards, the dataset from the download request can be accessed directly through a download key or using a API-link. We use a download key to import the dataset. This can be done using the script `1_2_GBIF_import.R`

### 1.3. GBIF data cleaning

Before looking into the different sources and "quality levels" of the data, in the dataset, we first clean the downloaded dataset: removing data points with flagged coordinates, temporal outliers, and low coordinate precision (\>1km). The full cleaning procedure is provided in more detail in the script `1_3_GBIF_data_cleaning.R`.

#### Procedure outline:

-   Visualize the raw data

-   Then, remove records with problematic coordinates and plot the flagged records.

-   Finally, we test for temporal outliers on the taxon level, and remove the records

To improve the precision and our confidence in the data further, we can use the metadata included in the dataset:

-   First, make a boxplot of the coordinate precisions reported for the insect data.

-   Then, calculate some summary stats for the coordinate uncertainties, make a table of each value's frequency, and inspect it.

-   Then, remove records with coordinate uncertainty \>3km and records with NA for coordinate uncertainty.

-   Save the cleaned files. The cleaned dataset is now called `insectdata_low_uncertainty`.

The cleaned data is visualized in @fig-coords, with points removed in the cleaning procedure marked in red.

```{r plot-cleaned}
#| label: fig-coords
#| layout-ncol: 1
#| fig-cap: "Coordinates for the cleaned occurrences (blue, $N=23434$) of B. Rhodani in Norway, and removed points (red, $N=124$)."
#| truncate: true
load(here("data", "source_data","insectdata.rda"))
load(here("data/derived_data/cleaned_insectdata.rda"))

#filter out the filtered-out observations
removed_obs <- subset(insectdata, !(occurrenceID %in% cleaned_insectdata$occurrenceID))

# make plot of raw v cleaned obs
ggplot() +
  coord_fixed() +
  borders(database="world",
          regions="Norway(?!:Svalbard)", colour = "gray70", fill = "gray90",
          xlim=c(4.5, 32)) +
  geom_point(data = cleaned_insectdata,
             aes(x = decimalLongitude, y = decimalLatitude),
             colour = "darkblue",
             size = 0.5) +
  geom_point(data = removed_obs,
             aes(x = decimalLongitude, y = decimalLatitude),
             colour = "red",
             size = 0.5) +
  coord_quickmap() +
  theme_classic()
```

### 1.4. Data exploration

The cleaned data originates from different data providers. The 10 largest data sources are shown below, and we see that after The University Museum (`NTNU-VM`), The Norwegian Environmental Agency (`miljodir`) and the Norwegian Institute for Nature Research (`NINA`) are the institutions providing the highest number of data points, as shown in @fig-institution-barplot.

```{r plot-institution-barplot}
#| label: fig-institution-barplot
#| layout-ncol: 1
#| fig-cap: "Number of registered occurences from different data sources."
#| truncate: true
###
# Load cleaned insectdata
load(here::here("data","derived_data","cleaned_insectdata.rda"))

# Summarize the number of occurrences by institution
df_institutions <- cleaned_insectdata %>%
  group_by(institutionCode) %>%
  summarize(N_occurrences = length(occurrenceID))%>%
  mutate(across(where(is.character), ~ na_if(.,""))) %>%
  filter(!is.na(institutionCode))

# Plot figure for top 10 institutions
df_institutions_barplot <- df_institutions %>% 
  arrange(desc(N_occurrences)) %>%
  slice(1:10) %>%
  ggplot(., aes(x = reorder(institutionCode,-N_occurrences), y = N_occurrences)) + 
  geom_bar(stat = "identity") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  xlab("Institution") + 
  ylab("Number of occurrences") +
  theme_classic()

df_institutions_barplot

```

As visualized in @fig-institutions-points, the *B. rhodani* data points are distributed across Norway, with higher concentrations of data points visible near large cities.

```{r plot-institutions}
#| label: fig-institutions-points
#| layout-ncol: 1
#| fig-cap: "Coordinates for the samples provided by the the largest data providers in the study: The NTNU University Museum (NTNU-VM), the Norwegian Environmental Agency (miljodir) and the Norwegian Institute for Nature Research (NINA)."
#| truncate: true
largest_providers <- cleaned_insectdata |> filter(institutionCode %in% c("NTNU-VM", "miljodir", "NINA"))
ggplot() +
  coord_fixed() +
  borders(database="world",
          regions="Norway(?!:Svalbard)", colour = "gray70", fill = "gray90",
          xlim=c(4.5, 32)) +
  geom_point(data = largest_providers,
             aes(x = decimalLongitude, 
                 y = decimalLatitude,
                 color = institutionCode),
             size = 0.5,
             alpha=0.5) +
  coord_quickmap() +
  facet_wrap(~institutionCode) +
  labs(color="Data provider") +
  theme_minimal() +
    theme(legend.position = "bottom") 
```

Another distinct feature of the data, is the fact that NTNU university museum samples seem to be heavily localized in middle / northern Norway, while the two other data sets are more evenly distributed across the country, with the exception of the northernmost Finnmark region, where the sampling effort is low for all data sets.

### 1.5. Creating presence-absence dataset

From inspecting the cleaned dataset, we see that both the Environmental Agency and NINA are missing information on the sampling protocol. By comparison, the NTNU University Museum provides reliable information about the sampling method for nearly all samples (14173 of 14184 occurrences), as well as individual counts, sampling duration- and area (mostly), and individual counts.

On the basis of this, we propose that the Environmental Agency and NINA data, together with the data from smaller providers, be treated as presence-only data. However, the NTNU University Museum dataset has enough structure for us to construct a presence-absence dataset, given comparable samples taken with the same protocol where *B. rhodani* is absent.

In order to find these, we can look at the `datasetKey` column in the dataset, and make a new GBIF request based on that. As it turns out, nearly all the samples from the NTNU Science Museum originate from a single dataset, the [Freshwater benthic invertebrates ecological collection](https://www.gbif.org/dataset/33591b80-0e31-480c-82ce-2f57211b10e6).

We created a new download request with the same parameters as in 1.1., except not restricted to *B. rhodani*, and filtered it to only contain samples obtained with sampling methods suitable for catching *B. rhodani*. Details on this procedure can be found in the script `1_5_GBIF_presence_absence_dataset.R`.

```{r plot-presence_absence}
#| label: fig-presence_absence
#| layout-ncol: 1
#| fig-cap: "Plot of presences/absences (1/0) of _B. Rhodani_ from the NTNU Freshwater benthic invertebrates ecological collection, filtered by samples taken with comparable sampling protocols."

# Load data
load("data/derived_data/presence_absence_dataset.Rda")

# plot
ggplot() +
    coord_fixed() +
    borders(database="world",
            regions="Norway(?!:Svalbard)", colour = "gray70", fill = "gray90",
            xlim=c(4.5, 32)) +
    geom_point(data = events_NTNU,
               aes(x = as.numeric(decimalLongitude), y = as.numeric(decimalLatitude),
               colour = as.factor(presence)),
               size = 0.5) + 
  labs(color="Presence", x="Longitude", y="Latitude") +
  coord_quickmap() + 
  theme_minimal()
```

## 2. WORLDCLIM environmental variables

We used the Worldclim bioclimatic variables to map the distribution of *B. rhodani.* Specifically, we used bioclim10 (Mean Temperature of The Warmest Quarter) and bioclim11 (Mean Temperature of The Coldest Quarter) at a 30 arc second resolution (\~ 1km\^2). Below, we plotted the distribution of the bio10 and bio11 values across Norway. These particular variables were chosen due to their assumed impact on water temperature, as water temperature is known to be important for *B. rhodani* development. Preliminary analysis revealed a significant but moderate correlation value between the two climate variables (r = 0.559, t(877822) = 631, p \< 0.001). Therefore, the two variables were used in the SDMs.

In this report, we chose to include WORLDCLIM 10 and 11 as predictor variables to our iSDM. In order to predict the distribution of *Baetis rhodani*, it could also be of interest to include land-cover data (i.e, CORINE land cover data, ref) or distance to rivers (i.e., created using the river network dataset provided by NVE, ref).

```{r plot worldclim}
#| label: fig-worldclim
#| layout-ncol: 1
#| fig-cap: "Plot of Mean Temperature of The Warmest Quarter and Mean Temperature of The Coldest Quarter across Norway"

load("data/source_data/wc2.1_30s/wc2.1_30s_bio_10.tif")
load("data/source_data/wc2.1_30s/wc2.1_30s_bio_11.tif")

# Convert rasters to data frames for ggplot
bio10_norway_df <- as.data.frame(bio10_norway, xy = TRUE)
bio11_norway_df <- as.data.frame(bio11_norway, xy = TRUE)

# Plot for bio10
bio10_plot <- ggplot(data = bio10_norway_df) +
  geom_raster(aes(x = x, y = y, fill = wc2.1_30s_bio_10)) + 
  borders(database = "world", regions = "Norway(?!:Svalbard)", colour = "gray70") +
  coord_quickmap() +
  labs(fill = "Mean Temperature of The Warmest Quarter (°C)")+
  theme_minimal() +
  theme(legend.position = "bottom")

# Plot for bio11
bio11_plot <- ggplot(data = bio11_norway_df) +
  geom_raster(aes(x = x, y = y, fill = wc2.1_30s_bio_11)) + 
  borders(database = "world", regions = "Norway(?!:Svalbard)", colour = "gray70") +
  coord_quickmap() +
  labs(fill = "Mean Temperature of The Coldest Quarter (°C)")+
  theme_minimal() +
  theme(legend.position = "bottom")

# Plot together
plot_grid(bio10_plot, bio11_plot)
```

## 3. Integrated Species Distribution Model

We set up the `INLA` raster to be used in the `pointedSDM` model according to the recommendations provided at (Mostert & O'Hara, 2023). Mesh coarseness was adjusted in order to balance performance and accuracy given the amount and spread of the data provided. All environmental variables were centered and scaled to mean zero and unit variance, and four different models were specified, according to the study questions outlined above.

For more details on model set up and mesh adjustment, see the script `5_integrated_SDM.R`.

# Results

Predictions of the likelihood of occurrence of B. rhodani were created for each of the four models, as seen in @fig-predictions.

```{r SDMs}
#| eval: true
#| echo: false
#| label: fig-predictions
#| layout-ncol: 2
#| fig-cap: "Predicted intensities of B. rhodani in Norway according to the fitted model"
#| fig-subcap: 
#| - Full presence-only
#| - Partial presence-only
#| - Presence-absence (NTNU-VM)
#| - Integrated model
#| truncate: true
# Run SDM script(s)
#source(here::here("code","5_integrated_SDM.R"))

### load models (new script?) ###
load("data/model_fits/po_full.rda")
load("data/model_fits/po_partial.rda")
load("data/model_fits/pa_only.rda")
load("data/model_fits/integrated.rda")

# Plot the prediction
plot(pred_po_full)
plot(pred_po_partial)
plot(pred_pa_only)
plot(pred_integrated)


```

```{r}
modelRun_po_full$summary.fixed
modelRun_po_full$summary.hyperpar
modelRun_po_partial$summary.fixed
modelRun_po_partial$summary.hyperpar
modelRun_pa_only$summary.fixed
modelRun_pa_only$summary.hyperpar
modelRun_integrated$summary.fixed
modelRun_integrated$summary.hyperpar


```

# Model interpretation and discussion

# Bibliography

::: {#refs}
:::

/li

Manual reference list

-   Alp, M., Keller, I., Westram, A. M., & Robinson, C. T. (2012). How river structure and biological traits influence gene flow: a population genetic study of two stream invertebrates with differing dispersal abilities. *Freshwater Biology*, *57*(5), 969--981. https://doi.org/10.1111/j.1365-2427.2012.02758.x

-   Artsdatabanken (2021). **Vanlig smådøgnflue *Baetis rhodani* (Pictet, 1843-45).** Retrieved 13.05.2024. <https://artsdatabanken.no/Taxon/Baetis%20rhodani/49481>

-   Bagge, P. (1995). Emergence and upstream flight of lotic mayflies and caddisflies (Ephemeroptera and Trichoptera) in a lake outlet, central Finland. *Entomologica Fennica*, *6*(2-3), 91--97. https://doi.org/10.33338/ef.83844

-   European Environment Agency (EEA) (2018). *CORINE Land Cover 2018 (CLC2018), Version 20_2018, European Environment Agency.* Retrieved from https://land.copernicus.eu/pan-european/corine-land-cover

-   GBIF.org (2024), GBIF Home Page. Available from: [https://www.gbif.or](https://www.gbif.org/)g.

-   Mostert, P. S., & O'Hara, R. B. (2023). PointedSDMs: An R package to help facilitate the construction of integrated species distribution models. *Methods in Ecology and Evolution*, 14, 1200--1207. <https://doi.org/10.1111/2041-210X.14091>

-   NVE (2020, October 7). Elvenettverk/ELVIS. Retrieved from: <https://www.nve.no/kart/kartdata/vassdragsdata/elvenettverk-elvis/>

-   Peredo Arce, A., Hörren, T., Schletterer, M., & Kail, J. (2021). How far can EPTs fly? A comparison of empirical flying distances of riverine invertebrates and existing dispersal metrics. *Ecological Indicators*, *125*, 107465. https://doi.org/10.1016/j.ecolind.2021.107465

Citing R and packages used, including package version.

```{r}
# Citing all packages used, including the package versions.
report::cite_packages()
```
