---
title: "main document"
format: html
editor: visual
---

## Info

This file documents the different steps in the mayfly case study, created for the course Advanced biology BI8091 spring 2024.

## Setup

Setup: Load needed packages, and download data.

```{r prep}
## Custom install & load function
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, repos = "http://cran.us.r-project.org")
  }
  require(x, character.only = TRUE)
}

package_vec <- c("here","dplyr","knitr","ggplot2","rgbif") # specify package

## executing install & load for each package
sapply(package_vec, install.load.package)

```

Download Baetis rhodani occurrence data from GBIF. Example can be found here "https://data-blog.gbif.org/post/downloading-long-species-lists-on-gbif/". Download criteria:

-   Species: Baetis rhodani

-   Geographic area: Norway

-   Basis of record: Everything

-   Year range: 1950-current

```{r data download}
# Load script
source(here::here("code","download.R"))


```

## Data cleaning

```{r}

```

## Data exploration

```{r}

```

## Analysis

```{r}

```

## Citations 

Citing R and packages used, including package version.

```{r}
# Citing all packages used, including the package versions.
report::cite_packages()
```
